Bandit Multiclass Linear Classification: \\ Efficient Algorithms for the Separable Case

### 2019-01-23 ICML-submission/mistake-lower-bound-for-ignorant-algorithms.tex
Clean-up Appendix I

\begin{algorithm}
\caption{\textsc{Adversary's strategy}}
\label{algorithm:adversary-strategy}
\textbf{Define} $T=N$ and $v_1, v_2, \dots, v_N$ as in Lemma~\ref{lemma:embed_d_gamma}.\\
\textbf{Define} $q_0=\frac{1}{\sqrt{T}}$. \\
\textbf{Initialize} $\textsc{phase}= 1$. \\
\For{$t=1,2,\dots,T$}{
    \If{$\textsc{phase}=1$}{
       \If{$p_t(1|v_t)\ge 1-q_0$}{
          $(x_t, y_t)\leftarrow (v_t, 1)$
        }
       \Else{
          $(x_t, y_t)\leftarrow (v_t, 2)$ \\
          $\textsc{phase}\leftarrow 2$
       }
    }
    \Else{
         $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$
    }
}
\end{algorithm}

Define the indicators
\begin{align*}
A_t &= \indicator{\forall \tau\le t, p_\tau(1|x_\tau)<1-q_0} \\
B_t &= \indicator{\exists \tau\le t, p_\tau(1|x_\tau)\ge 1-q_0
 \text{\ and\ } \forall s\in[\tau,t), \widehat y_s \neq y_s}.
\end{align*}
Then we have
\begin{align}
& \mathbf{E}\left[\sum_{t=1}^{T} \indicator{\widehat y_t\neq y_t}\right] \nonumber \\
& \ge \mathbf{E}\left[\sum_{t=1}^T \indicator{\widehat y_t\neq y_t}A_t\right] \nonumber \\
& \qquad + \mathbf{E}\left[\sum_{t=1}^T \indicator{\widehat y_t\neq y_t}B_t\right] \nonumber \\
& = \mathbf{E}\left[\sum_{t=1}^T \mathbf{E}_t\left[\indicator{\widehat y_t\neq y_t}A_t\right]\right] + \mathbf{E}\left[\sum_{t=1}^T B_{t+1}\right] \nonumber \\
& \ge \mathbf{E}\left[\sum_{t=1}^T q_0 A_t\right] + \mathbf{E}\left[\sum_{t=1}^T B_{t+1}\right],
\label{equation:mistake_lower_bound_temp}
\end{align}
where in the first inequality we use the fact that $A_t$ and $B_t$ cannot be $1$
simultaneously; in the equality, we use $B_{t+1}=B_t\indicator{\widehat y_t\neq
y_t}$ by $B_t$'s definition; in the last inequality we use the fact that when
$t\le N$ and $A_t=1$, it must be $y_t=1$ and $\Pr[\widehat
y_t=y_t]=p_t(1|x_t)<1-q_0$. Now, denote $T_1 = \argmin_{\tau} \{
p_\tau(1|x_\tau) \ge 1-q_0 \}$ (if such $\tau$ does not exist or is larger than
$T$, we simply let $T_1=T+1$). Then $A_t=1$ for all $t\le
T_1-1$.

Note that $B_{T_1}=1$, and for $t \ge T_1$ the adversary will switch to
``$\textsc{phase}=2$'' and has $y_t=2$. Therefore,
$\Exp[B_{t+1}=0|B_{t}=1]=\Pr[\widehat y_t=y_t~|~B_t=1] = \Pr[\widehat
y_t=2|B_t=1].$ Note that when $B_t=1$, by definition there exists a $\tau\le t$
with $p_\tau(1|x_\tau)\ge 1-q_0$, and the algorithm all makes mistakes at $\tau,
\tau+1, \ldots, t$. Since the algorithm is ignorant, and when
$\textsc{phase}=2$, the features all remain the same, we have
$p_t(1|x_t)=p_{t-1}(1|x_{t-1})=\cdots=p_\tau(1|x_\tau)\ge 1-q_0$. Thus we can
further upper bound $\Exp[B_{t+1}=0|B_{t}=1]$ by $\Pr[\widehat y_t=2|B_t=1]\leq
q_0$. Thus the second term in \eqref{equation:mistake_lower_bound_temp} can be
calculated as

\begin{align*}
\mathbf{E}\left[\sum_{t=1}^T B_{t+1}\right]
& \ge \mathbf{E}\left[\sum_{t=T_1}^T (1-q_0)^{t-T_1+1} \right] \\
& = \frac{1-q_0}{q_0}\mathbf{E}\left[1-(1-q_0)^{T - T_1 + 1}\right] \; .
\end{align*}
Combining \eqref{equation:mistake_lower_bound_temp} and the above inequalities, we get
\begin{multline}
\label{equation:mistake_lower_bound_temp2}
\mathbf{E} \left[\sum_{t=1}^T \indicator{\widehat y_t\neq y_t}\right] \ge
\\ \mathbf{E}\left[q_0(T_1-1)+\frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-T_1+1}\right)\right].
\end{multline}
In the case $T_1\ge \frac{1}{2}T + 1$, the right-hand side of
\eqref{equation:mistake_lower_bound_temp2} is lower bounded by
$\frac{1}{2}q_0 T = \frac{1}{2} \sqrt{T}$. In the case
$T_1< \frac{1}{2}T+1$, it is lower bounded by
\begin{align*}
& \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2}T}\right) \\
& = \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2q_0^2}}\right) \\
& \ge \frac{1-\frac{1}{\sqrt{2}}}{q_0}\left(1-\frac{1}{\sqrt{e}}\right) \\
& \ge \frac{1}{10} \sqrt{T} \; .
\end{align*}

Observe that in phase 1, the labels are equal to $1$ and in phase 2 the labels
are equal to $2$. Let $t$ be the round in which the switch happens. (If it never
happens, define $t=T+1$.) Note that $(x_t, y_t)=(x_{t+1}, y_{t+1})= \dots =
(x_T, y_T) = (v_t, 2)$. Consider the vectors $u_1, u_2, \dots, u_N$ as defined
in Lemma~\ref{lemma:embed_d_gamma} We claim that $w_1=-u_t/2$ and $w_2=u_t/2$
satisfy the conditions of strong linear separabaility.

Clearly $\norm{w_1}^2 + \norm{w_2}^2 \le (\norm{w_1} + \norm{w_2})^2 \le
(\frac{1}{2} + \frac{1}{2})^2 \le 1$. By Lemma~\ref{lemma:embed_d_gamma}, we
have $\ip{w_2/2}{x_s} = \ip{u_t/2}{v_t} \ge \gamma/2, \forall s \ge t$ and
$\ip{w_2/2}{x_s} = \ip{u_t/2}{v_s} \le - \gamma/2$ for all $s<t$. Similarly,
$\ip{w_1/2}{x_s} \le -\gamma/2$ for all $s \ge t$ and $\ip{w_1/2}{x_s} \ge
\gamma/2$ for all $s < t$. Thus, the examples are strongly separable with margin
$\gamma$.
\end{proof}

>>>>

\begin{algorithm}
\caption{\textsc{Adversary's strategy}}
\label{algorithm:adversary-strategy}
\textbf{Define} $T=N$ and $v_1, v_2, \dots, v_N$ as in Lemma~\ref{lemma:embed_d_gamma}.\\
\textbf{Define} $q_0=\frac{1}{\sqrt{T}}$. \\
\textbf{Initialize} $\textsc{phase}= 1$. \\
\For{$t=1,2,\dots,T$}{
    \If{$\textsc{phase}=1$}{
       \If{$p_t(1|v_t) < 1-q_0$}{
          $(x_t, y_t)\leftarrow (v_t, 1)$
        }
       \Else{
          $(x_t, y_t)\leftarrow (v_t, 2)$ \\
          $\textsc{phase}\leftarrow 2$
       }
    }
    \Else{
         $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$
    }
}
\end{algorithm}

Let $\tau$ be the time step $t$ in which the adversary sets $\textsc{phase}\leftarrow 2$.
If the adversary never sets $\textsc{phase}\leftarrow 2$, we define $\tau = T + 1$.
Then,
\begin{align}
& \Exp \left[\sum_{t=1}^T \indicator{\widehat y_t\neq y_t}\right] \nonumber \\
& \ge \Exp\left[\sum_{t=1}^{\tau - 1} \indicator{\widehat y_t\neq y_t}\right]
+ \Exp\left[\sum_{t=\tau}^T \indicator{\widehat y_t\neq y_t}\right] \; .
\end{align}
We upper bound each of last two terms separately.

In rounds $1,2,\dots,\tau-1$, the algorithm predicts the incorrect class $2$
with probability at least $q_0$. Thus,
$$
\Exp\left[\sum_{t=1}^{\tau - 1} \indicator{\widehat y_t\neq y_t}\right] = q_0 \Exp[\tau] \; .
$$
In rounds $\tau, \tau+1, \dots, T$, all the examples are the same and are equal
to $(v_\tau, 2)$. Let $s$ be the first time step $t$ such that $t \ge \tau$
and the algorithm makes a correct prediction. If the algorithm makes mistakes
in all rounds $\tau, \tau+1, \dots, T$, we define $s = T+1$.
By definition the algorithm makes mistakes in rounds $\tau, \tau+1, \dots, s-1$
Then,
$$
\Exp\left[\sum_{t=\tau}^T \indicator{\widehat y_t\neq y_t}\right] \ge \Exp[s-\tau].
$$
Since the algorithm is ignorant, conditioned on $\tau$, the random variable $s-\tau$
has a truncated geometric distribution with parameter $p_\tau(2|v_\tau) \le q_0$.
Therefore,
$$
\Exp[s-\tau ~|~ \tau]
\ge
\Exp \left[\frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-\tau+1}\right) ~\middle|~ \tau \right] \; .
$$

Combining the two, we have that
\begin{align*}
& \Exp \left[\sum_{t=1}^T \indicator{\widehat y_t\neq y_t}\right] \nonumber \\
& \ge q_0 \Exp[\tau] + \Exp \left[\frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-T_1+1}\right)\right] \\
& =  \Exp \left[ \Exp \left[ q_0 \tau + \frac{1-q_0}{q_0}\left(1-(1-q_0)^{T-\tau+1}\right) ~\middle|~ \tau \right] \right] \; .
\end{align*}
We lower bound the last expression by considering two cases for $\tau$.
If $\tau \ge \frac{1}{2}T + 1$, then the last expression is lower bounded by
$\frac{1}{2}q_0 T = \frac{1}{2} \sqrt{T}$. If
$\tau < \frac{1}{2}T+1$, it is lower bounded by
\begin{align*}
& \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2}T}\right) \\
& = \frac{1-q_0}{q_0}\left(1-(1-q_0)^{\frac{1}{2q_0^2}}\right) \\
& \ge \frac{1-\frac{1}{\sqrt{2}}}{q_0}\left(1-\frac{1}{\sqrt{e}}\right) \\
& \ge \frac{1}{10} \sqrt{T} \; .
\end{align*}

Observe that in phase 1, the labels are equal to $1$ and in phase 2 the labels
are equal to $2$. Note that $(x_\tau, y_\tau)=(x_{\tau+1}, y_{\tau+1})= \dots =
(x_T, y_T) = (v_t, 2)$. Consider the vectors $u_1, u_2, \dots, u_N$ as defined
in Lemma~\ref{lemma:embed_d_gamma} We claim that $w_1=-u_\tau/2$ and $w_2=u_\tau/2$
satisfy the conditions of strong linear separabaility.

Clearly $\norm{w_1}^2 + \norm{w_2}^2 \le (\norm{w_1} + \norm{w_2})^2 \le
(\frac{1}{2} + \frac{1}{2})^2 \le 1$. By Lemma~\ref{lemma:embed_d_gamma}, we
have $\ip{w_2/2}{x_t} = \ip{u_\tau/2}{v_\tau} \ge \gamma/2, \forall t \ge \tau$ and
$\ip{w_2/2}{x_t} = \ip{u_\tau/2}{v_t} \le - \gamma/2$ for all $t < \tau$. Similarly,
$\ip{w_1/2}{x_t} \le -\gamma/2$ for all $t \ge \tau$ and $\ip{w_1/2}{x_t} \ge
\gamma/2$ for all $t < \tau$. Thus, the examples are strongly separable with margin
$\gamma$.
\end{proof}
